---
title: "Exercise 2_1: Import Vector Data"
subtitle: 'Geodata & Spatial Regression'
author: Anne-Kathrin Stroppe, Dennis Abel
format: 
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
editor: visual
callout-icon: false
callout-appearance: minimal
---

### Exercises

Let's get started with some hands-on exercises and exploring some data!
This exercise focuses on importing different vector data geometries from various sources and getting used to simple feature data tables.

In the folder `data`, you can find the data files prepped for all the exercises which rely on preexisting data sets.
However, if you like to play around with your data, feel free to do so!

::: callout-note
## 🏋 Exercise 1

Load the administrative borders of Germany.
For this purpose, import the shapefile `VG250_STA.shp` from the respective folder and assign it to an object named `germany`.

Plot the newly created object.
:::

::: {.callout-caution collapse="true"}
## 💡 Tip

Don't forget to load the packages `sf` and set your working directory before starting this exercise.
:::

The `R` console output already gives you some information on the data you just loaded.
You also see that the object `germany` appeared in your environment.
Let's explore the data set a little bit closely.

::: callout-note
## 🏋 Exercise 2

Explore the data and answer the following questions:

-   What is the geometry type?
-   What is the ID of the coordinate reference system?
-   What is the name of the column containing the information on geometries?
-   Which attributes are assigned to the geometries?
:::

::: {.callout-caution collapse="true"}
## 💡 Tip

You can either extract the information directly from the output or problematically with functions from base `R` and the `sf` package (you may want to have a look, e.g., at the help files `?sf::st_geometry` or `sf::st_crs()`).
:::

::: callout-note
## 🏋 Exercise 3

Do you have an idea why there are 11 observations instead of only 1, even though we expected a shapefile with the polygon of Germany?
Create a new object (`germany_new)`that contains only one observation filtered by the variable `GF == 4`.
The data provider, the Federal Agency for Cartography and Geodesy (BKG), has named this variable, which stands for "Geofactor" (German: "Geofaktor").
Maybe you can figure out what this category means after plotting the data.
:::

::: {.callout-caution collapse="true"}
## 💡 Tip

You can apply standard `R` procedures to filter data, such as `dplry`'s `filter()` command.
:::

::: callout-note
## 🏋 Exercise 4

CRS handling is a common source of errors.
There are two functions that sound similar but do very different things:

-   `sf::st_set_crs()` — **relabels** the CRS without changing any coordinate values
-   `sf::st_transform()` — actually **recalculates** the coordinates into a new CRS

Here is a point representing Berlin in geographic coordinates (longitude/latitude, WGS84/EPSG:4326):

```{r}
#| eval: false
berlin <- sf::st_sfc(sf::st_point(c(13.405, 52.52)), crs = 4326)
```

- `germany_new` is stored in EPSG:25832 (a projected CRS, units in metres).
    Use `sf::st_set_crs()` to *force* `berlin`'s CRS to match `germany_new`'s CRS without transforming.
    Plot `germany_new` and add this re-labeled point on top

-  Now use `sf::st_transform()` on the *original* `berlin` point to properly convert it to `germany_new`'s CRS, and plot again.
:::

::: {.callout-caution collapse="true"}
## 💡 Tip

For Plotting use

```{r}
#| eval: false

plot(germany_new)
plot(berlin_wrong, add = TRUE, col = "red", pch = 19, cex = 2)
```


:::

### Solutions

::: {.callout-tip collapse="true"}
## ✅ Solution 1

```{r}
#| eval: false
library(sf)

germany <- sf::read_sf("./data/VG250_STA.shp") 

plot(germany)
```

```{r}
#| echo: false
library(sf)

germany <- sf::read_sf("../../data/VG250_STA.shp") 

plot(germany)
```
:::

::: {.callout-tip collapse="true"}
## ✅ Solution 2

```{r}
sf::st_geometry(germany) # Multipolygon: Set of several Polygons. Each row is a polygon.

attr(germany, "sf_column") # geometry

sf::st_crs(germany) # ETRS89, ID = EPSG:25832

names(germany) # Name, Area Type, Area Size, Population Size ,...
```
:::

::: {.callout-tip collapse="true"}
## ✅ Solution 3

```{r}
# It seems like the shapefile contains not only Germany's land area but also coastal areas and lakes.
# filter 
germany_new <-
  germany |>
  dplyr::filter(GF == 4)

plot(germany_new)
```
:::

::: {.callout-tip collapse="true"}
## ✅ Solution 4

```{r}

berlin <- sf::st_sfc(sf::st_point(c(13.405, 52.52)), crs = 4326)

# a) Force the (wrong!) CRS of germany_new onto berlin
berlin_wrong <- sf::st_set_crs(berlin, sf::st_crs(germany_new))

plot(germany_new$geometry)
plot(berlin_wrong, add = TRUE, col = "red", pch = 19, cex = 2)
```


```{r}
# b) Properly transform the original berlin point instead
berlin_correct <- sf::st_transform(berlin, crs=25832)

plot(germany_new$geometry)
plot(berlin_correct, add = TRUE, col = "red", pch = 19, cex = 2)
```



```{r}
# c) Compare the bounding boxes
sf::st_bbox(berlin_wrong)
sf::st_bbox(berlin_correct)
```


:::
