---
title: "Exercise 4_2: Subsetting and Linking"
subtitle: 'Geodata & Spatial Regression'
author: "Anne-Kathrin Stroppe, Dennis Abel"
format: 
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
editor: visual
callout-icon: false
callout-appearance: minimal
editor_options: 
  chunk_output_type: console
---

### Exercises

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 1

In this exercise, you will combine administrative boundary data with OpenStreetMap point data to explore the spatial distribution of dentists and veterinarians in Cologne and its surrounding districts. 

First, load and prepare the German district shapefile, transform it to an appropriate projected coordinate reference system, and extract Cologne (`AGS == "05315"`) together with its neighbouring districts. 

Then, use the bounding box of this study area (in WGS84!) to query OpenStreetMap for amenities tagged as dentists or veterinary practices. After cleaning the retrieved data and retaining only point geometries with valid amenity information, transform the OSM data to match the CRS of the district boundaries. Finally, create a map with `ggplot2` that shows the Cologne region and its surrounding districts, with dentists and veterinarians displayed as color-coded point features.
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

Pay attention to the different spatial data formats and coordinate reference systems used in this workflow. The district boundaries are read from a shapefile, while the amenity data are downloaded from OpenStreetMap and may contain several geometry types, such as points, lines, and polygons. For this exercise, you only need the `osm_points` object. Make sure to remove missing values in the amenity variable before plotting, and always transform the OSM point data to the same CRS as the district polygons before combining both layers in one `ggplot()` map. You need the function `sf::st_transform()` to change the CRS.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 2

In the second part of the exercise, you will move from mapping individual dentists and veterinarians to comparing their relative availability across Cologne and its surrounding districts. 

To do this, spatially join the point data to the district polygons so that each amenity point receives the district identifier of the polygon in which it is located. 

Then, count the number of dentists and veterinarians separately for each district and combine these counts with district-level population data. 

Use the population values from our district-attribute table to calculate amenity densities per 100,000 inhabitants, which allows comparison between districts of different population sizes. 

Finally, reshape the resulting table so that the density values can be attached back to the district geometries, and create choropleth maps showing the spatial variation in dentist and veterinary densities across the study area.
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

The main challenge in this task is to move between spatial and tabular operations in the correct order. Use `st_join()` with st_within to assign each point to the district polygon that contains it. Use `st_drop_geometry()` when you no longer need the point geometries and want to perform ordinary table operations. Use `filter()` to remove points that were not matched to any district, and `count()` to calculate the number of amenities by district and amenity type. Use `pivot_wider()` to create one density column for dentists and one for veterinarians before joining the results back to the district shapefile. 
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 3

To wrap up, let's do some stuff with other data formats. Sample 50 points that fall within the boundaries of the city of Cologne.
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

Sampling is straightforward: Apply the `sf::st_sample` to the Cologne vector data, but make sure to apply the `sf::st_as_sf()` function afterward to receive a full-fledged data table (with a geometry column only).
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 4

Create a new raster layer comprising Cologne's immigrant rates based on the raster layers from the previous exercises. Extract the immigrant rate value at each position of the previously sampled points as a vector. What is your observation?
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

You need the `citizens_cologne.tif` and `inhabitants_cologne.tif` files in the `./data` folder. Due to severe data protection measures, the German Census 2012 data could be quite sparse.
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 5

Use an adequate method of raster extraction to gather information in the geographic surroundings of a point. What is your observation now?
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

Assume that people move in a 1,000-meter radius around their location. Thus, extracting information on buffers of 1,000 meters around the points might be interesting using the option `sf::st_buffer()` function. In that case, you should also set a descriptive statistics function, e.g., with the option `fun = mean` and its helpful companion option to consider missing values `na.rm = TRUE`.
:::

### Solutions

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 1

```{r}
#| eval: false
# load packages
library(sf)
library(dplyr)
library(ggplot2)
library(patchwork)

# Import data
german_districts <- 
  sf::read_sf("./data/VG250_KRS.shp") |>  
  sf::st_transform(3035) |>  
  dplyr::select(AGS, GEN)

# filter Cologne
cologne <-
  german_districts |> 
  dplyr::filter(AGS == "05315")

# filter surrounding districts and append with Cologne data
cologne_sur <-
  german_districts |> 
  dplyr::filter(lengths(sf::st_touches(german_districts, cologne)) > 0) |> 
  dplyr::bind_rows(cologne) 

# Generate a bounding box for Cologne surroundings area and transform to WGS84
bbox <- sf::st_bbox(cologne_sur) |> 
  sf::st_transform(crs = 4326)

# Access OSM data and retrieve all veterinarians for your region
osmdata::set_overpass_url("https://overpass-api.de/api/interpreter")

dents_vets <-
  osmdata::opq(bbox = bbox, timeout = 60) |> 
  osmdata::add_osm_feature(
    key = "amenity", 
    value = c("dentist", "veterinary")
  ) |> 
  osmdata::osmdata_sf()

dents_vets
# Some larger facilities are probably stored as polygons and not points
# For this exercise, we ignore these but keep in mind that the same feature
# might be stored in different vector types.

# Store only the points vector and keep only the amenity variable and geometry, which
# stores whether the point represents a dentist or a veterinarian.
dents_vets <- dents_vets$osm_points |> 
  select(amenity)

# The data also contains a few NAs. Filter these.
dents_vets <- dents_vets |> 
  filter(!is.na(amenity))

# Plot the points in color-coding for both type and add the borders from the
# Cologne and neighbors shapefile. Keep in mind to align the CRS of both vector files
# before it.
dents_vets <-
  sf::st_transform(
    dents_vets, 
    crs = sf::st_crs(cologne_sur)
  )
  
ggplot() +
  geom_sf(data = cologne_sur, fill = "grey95", color = "black") +
  geom_sf(data = dents_vets, aes(color = amenity), size = 1, alpha = .5) +
  theme_minimal()

```

```{r}
#| echo: false
# load packages
library(sf)
library(dplyr)
library(ggplot2)
library(patchwork)

# Import data
german_districts <- 
  sf::read_sf("../../data/VG250_KRS.shp") |>  
  sf::st_transform(3035) |>  
  dplyr::select(AGS, GEN)

# filter Cologne
cologne <-
  german_districts |> 
  dplyr::filter(AGS == "05315")

# filter surrounding districts and append with Cologne data
cologne_sur <-
  german_districts |> 
  dplyr::filter(lengths(sf::st_touches(german_districts, cologne)) > 0) |> 
  dplyr::bind_rows(cologne) 

# Generate a bounding box for Cologne surroundings area and transform to WGS84
bbox <- sf::st_bbox(cologne_sur) |> 
  sf::st_transform(crs = 4326)

# Access OSM data and retrieve all veterinarians for your region
osmdata::set_overpass_url("https://overpass-api.de/api/interpreter")

dents_vets <-
  osmdata::opq(bbox = bbox, timeout = 60) |> 
  osmdata::add_osm_feature(
    key = "amenity", 
    value = c("dentist", "veterinary")
  ) |> 
  osmdata::osmdata_sf()

dents_vets
# Some larger facilities are probably stored as polygons and not points
# For this exercise, we ignore these but keep in mind that the same feature
# might be stored in different vector types.

# Store only the points vector and keep only the amenity variable and geometry, which
# stores whether the point represents a dentist or a veterinarian.
dents_vets <- dents_vets$osm_points |> 
  select(amenity)

# The data also contains a few NAs. Filter these.
dents_vets <- dents_vets |> 
  filter(!is.na(amenity))

# Plot the points in color-coding for both type and add the borders from the
# Cologne and neighbors shapefile. Keep in mind to align the CRS of both vector files
# before it.
dents_vets <-
  sf::st_transform(
    dents_vets, 
    crs = sf::st_crs(cologne_sur)
  )
  
ggplot() +
  geom_sf(data = cologne_sur, fill = "grey95", color = "black") +
  geom_sf(data = dents_vets, aes(color = amenity), size = 1, alpha = .5) +
  theme_minimal()

```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 2

```{r}
#| eval: false
# we want to understand the density of dentists and veterinarians for each of
# the seven districts. Count the number of points within each polygon (by type)
# and calculate as share of the population.

# We start by joining the districts to the points, which gives each point the
# district ID
points_with_AGS <- st_join(
  dents_vets,
  cologne_sur,
  join = st_within
)

points_with_AGS

# Count points by polygon and amenity type
point_counts <- points_with_AGS  |> 
  st_drop_geometry() |> 
  filter(!is.na(AGS)) |> 
  count(AGS, amenity, name = "n")

point_counts

# Add population data from attribute table and calculate density by 100000 inhabitants
attributes_districts <- 
  readr::read_delim("./data/attributes_districts.csv", delim = ";") |> 
  select(population, AGS)

point_counts <- 
  dplyr::left_join(point_counts, attributes_districts,
                   by = "AGS") |> 
  dplyr::mutate(
    dens = (n*100000)/population
  )

point_counts

# Convert to wide format and attach to shapefiles
point_counts_wide <- point_counts |> 
  select(!c(n, population)) |> 
  tidyr::pivot_wider(
    names_from = amenity,
    values_from = dens,
    values_fill = 0,
    names_prefix = "dens_"
  )

point_counts_wide

# Attach your density values to your districts shapefile and plot both variables.
cologne_sur <- cologne_sur |> 
  left_join(point_counts_wide, by = "AGS")

dentist_plot <- ggplot(cologne_sur) +
  geom_sf(aes(fill = dens_dentist)) +
  scale_fill_viridis_c() +
  theme_minimal()

veterinary_plot <- ggplot(cologne_sur) +
  geom_sf(aes(fill = dens_veterinary)) +
  scale_fill_viridis_c() +
  theme_minimal()

dentist_plot | veterinary_plot

cologne_sur_long <- cologne_sur |> 
  tidyr::pivot_longer(
    cols = starts_with("dens_"),
    names_to = "amenity",
    values_to = "density"
  )

ggplot(cologne_sur_long) +
  geom_sf(aes(fill = density)) +
  facet_wrap(~ amenity) +
  theme_minimal()
```

```{r}
#| echo: false
# we want to understand the density of dentists and veterinarians for each of
# the seven districts. Count the number of points within each polygon (by type)
# and calculate as share of the population.

# We start by joining the districts to the points, which gives each point the
# district ID
points_with_AGS <- st_join(
  dents_vets,
  cologne_sur,
  join = st_within
)

points_with_AGS

# Count points by polygon and amenity type
point_counts <- points_with_AGS  |> 
  st_drop_geometry() |> 
  filter(!is.na(AGS)) |> 
  count(AGS, amenity, name = "n")

point_counts

# Add population data from attribute table and calculate density by 100000 inhabitants
attributes_districts <- 
  readr::read_delim("../../data/attributes_districts.csv", delim = ";") |> 
  select(population, AGS)

point_counts <- 
  dplyr::left_join(point_counts, attributes_districts,
                   by = "AGS") |> 
  dplyr::mutate(
    dens = (n*100000)/population
  )

point_counts

# Convert to wide format and attach to shapefiles
point_counts_wide <- point_counts |> 
  select(!c(n, population)) |> 
  tidyr::pivot_wider(
    names_from = amenity,
    values_from = dens,
    values_fill = 0,
    names_prefix = "dens_"
  )

point_counts_wide

# Attach your density values to your districts shapefile and plot both variables.
cologne_sur <- cologne_sur |> 
  left_join(point_counts_wide, by = "AGS")

dentist_plot <- ggplot(cologne_sur) +
  geom_sf(aes(fill = dens_dentist)) +
  scale_fill_viridis_c() +
  theme_minimal()

veterinary_plot <- ggplot(cologne_sur) +
  geom_sf(aes(fill = dens_veterinary)) +
  scale_fill_viridis_c() +
  theme_minimal()

dentist_plot | veterinary_plot

cologne_sur_long <- cologne_sur |> 
  tidyr::pivot_longer(
    cols = starts_with("dens_"),
    names_to = "amenity",
    values_to = "density"
  )

ggplot(cologne_sur_long) +
  geom_sf(aes(fill = density)) +
  facet_wrap(~ amenity) +
  theme_minimal()
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 3

```{r}
cologne_50_points <-
  cologne |>  
  sf::st_sample(50) |> 
  sf::st_as_sf()
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 4

```{r}
#| eval: false
library(terra)

# load all layers
citizens_cologne <- terra::rast("./data/citizens_cologne.tif")
inhabitants_cologne <- terra::rast("./data/inhabitants_cologne.tif")

# define missing values
citizens_cologne[citizens_cologne == -9] <- NA
inhabitants_cologne[inhabitants_cologne == -9] <- NA

# create immigrants layer
immigrants_cologne <- inhabitants_cologne - citizens_cologne

# something's off - let's get rid of the wierd negative numbers
immigrants_cologne[immigrants_cologne < 0] <- NA

immigrant_rates <-
  immigrants_cologne * 100 / inhabitants_cologne

immigrant_rates_at_point <- terra::extract(immigrant_rates, cologne_50_points)

immigrant_rates_at_point

# There are a lot of missing values.
```

```{r}
#| echo: false
library(terra)

# load all layers
citizens_cologne <- terra::rast("../../data/citizens_cologne.tif")
inhabitants_cologne <- terra::rast("../../data/inhabitants_cologne.tif")

# define missing values
citizens_cologne[citizens_cologne == -9] <- NA
inhabitants_cologne[inhabitants_cologne == -9] <- NA

# create immigrants layer
immigrants_cologne <- inhabitants_cologne - citizens_cologne

# something's off - let's get rid of the wierd negative numbers
immigrants_cologne[immigrants_cologne < 0] <- NA

immigrant_rates <-
  immigrants_cologne * 100 / inhabitants_cologne

immigrant_rates_at_point <- terra::extract(immigrant_rates, cologne_50_points)

immigrant_rates_at_point

# There are a lot of missing values.
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 5

```{r}
immigrant_rates_1000m_buffer <-
  terra::extract(
    immigrant_rates, 
    sf::st_buffer(cologne_50_points, 1000), 
    fun = mean,
    na.rm = TRUE
    )

immigrant_rates_1000m_buffer
```
:::
