---
title: "Exercise 5_1: Geocoding & Proximity"
subtitle: 'Geodata & Spatial Regression'
author: Anne Stroppe & Dennis Abel
format:
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
editor: visual
callout-icon: false
callout-appearance: minimal
editor_options:
  chunk_output_type: console
---

### Exercises

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 1: Geocoding

Load a number of hospitals from the hospital registry (`./data/krankenhausverzeichnis.xlsx`) or choose any other list of addresses.

1.  Geocode a set of addresses using `tidygeocoder::geocode()` with `method = "osm"`. How many were matched successfully?
2.  Convert the result to an `sf` point layer (CRS 4326), then transform to CRS 3035.
3.  Map the geocoded hospitals.
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 1

```{r}
#| eval: false

# load hospital data
hospitals_cologne <- readxl::read_excel("data/krankenhausverzeichnis.xlsx", sheet = 5, skip = 2, col_names=TRUE) |>
  dplyr::select(1:2,4:9) |>
  setNames(c("state", "district", "name", "name2", "street", "housenumber", "zip", "place")) |>
  dplyr::mutate(address = paste(street, housenumber, zip, place, "Germany")) 
```
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 2

Use `sf::st_as_sf(coords = c("long", "lat"), crs = 4326)` to convert.
Note the order: longitude first, then latitude.
Check for `NA` coordinates before converting with `dplyr::filter(!is.na(long))`.
:::

:::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 2: Distance & Nearest Neighbour

Using the Cologne voting precincts (`Stimmbezirk.shp`) and a point dataset of your choice, compute the distance from each precinct centroid to the nearest point feature.
Map the result.

**Choose your point dataset — pick whichever fits your interests:**

::: panel-tabset
### Option A: Hospitals (from Exercise 1)

Use the hospitals you already geocoded.

```{r}
#| eval: false
# Use the hospitals_sf object from Exercise 1
hospitals_sf_3035 <- sf::st_transform(hospitals_sf, 3035)
```

### Option B: NRW train stops (data folder)

Use the train stop data already in the `./data` folder.
Note: CRS is missing — set it manually.

```{r}
#| eval: false
train_stops <- sf::read_sf("./data/nrw_pt_osmtrainstops.shp") |>
  sf::st_set_crs(25832) |>
  sf::st_transform(3035)   # coordinates already in LAEA Europe
```

###  Option C: Your own OSM feature

Query any feature from OpenStreetMap using the `osmdata` package (as in Session 4).
Examples: pharmacies, schools, supermarkets, parks, bus stops.

```{r}
#| eval: false
library(osmdata)

your_features <- osmdata::getbb("Köln") |>
  osmdata::opq(timeout = 25) |>
  osmdata::add_osm_feature(
    key   = "amenity",       # change key/value to your feature of interest
    value = "pharmacy"       # e.g. "school", "supermarket", "post_office"
  ) |>
  osmdata::osmdata_sf() |>
  _$osm_points |>
  sf::st_transform(3035)
```
:::

::::


::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

**Steps for all options:**

1.  Load `Stimmbezirk.shp`, set CRS to 3035, compute centroids with `sf::st_centroid()`
2.  Find the nearest feature for each centroid using `sf::st_nearest_feature()`
3.  Compute distances with `sf::st_distance(..., by_element = TRUE)`
4.  Map the result: Which precincts have the worst access?
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

Remember: `st_nearest_feature()` returns an *index*, not a distance.
Use that index to subset your point layer before calling `st_distance(..., by_element = TRUE)`.
Convert the result to numeric metres with `as.numeric()`.
:::


::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 3: Buffers & Hospital Coverage

So far, we have measured hospital accessibility as the straight-line distance from each polling district's centroid to the nearest hospital. This is a useful first approximation but how convincing is it really?
Consider what a centroid represents: a single point summarising an entire area. A polling district might have its centroid close to a hospital, yet most of its residents could live at the far end of the district, far from any facility. 
The centroid distance measure implicitly assumes that all residents live at that one point but this assumption becomes increasingly problematic as districts grow larger or more irregularly shaped.

Instead, we are interes in the question: **What share of each polling district's population lives within reach of a hospital?**

In Session 4, we used `sf::st_buffer()` to create circular buffers around points and `terra::extract()` to summarize raster values within them.
Let's apply the same idea here: instead of measuring *distance* to the nearest hospital.

::: panel-tabset
# Workflow

1.  Create a 1,000m buffer around each hospital in `hospitals_sf` using `sf::st_buffer()`.
2.  Intersect the Cologne `stimmbezirke` (from Exercise 2) with these buffers to get the part of each precinct that lies within 1,000m of a hospital.
3.  Load `./data/inhabitants_cologne.tif` and use `terra::extract()` to get (a) the total population per Stimmbezirk and (b) the population within the "covered" part(s).
4.  Compute the share (in %) of each Stimmbezirk's population that lives within 1,000m of a hospital, and map the result.
:::
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 1

```{r}
#| eval: false

hospital_buffers <- sf::st_buffer(hospitals_sf, dist = 1000)
```
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 2

`sf::st_intersection(x, y)` returns only the parts of `x`'s geometries that overlap with `y`, keeping `x`'s attributes — unlike `sf::st_join()`, the geometries themselves get cut.

```{r}
#| eval: false
stimmbezirke_covered <- sf::st_intersection(stimmbezirke, hospital_buffers)
```

Reminder: If a Stimmbezirk overlaps more than one hospital buffer, it appears as multiple rows in the result.

:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 3

`terra::extract(raster, polygons, fun = sum, na.rm = TRUE)` returns a data frame with an `ID` column and one value column. Use the second column (`[, 2]`).

The raster uses `-9` as a placeholder for missing values; replace it with `NA` first:

```{r}
#| eval: false
inhabitants_cologne <- terra::rast("./data/inhabitants_cologne.tif") |>
  terra::subst(from = -9, to = NA)
```

Make sure `stimmbezirke` and `hospital_buffers` are transformed to the raster's CRS (`terra::crs(inhabitants_cologne)`) before extracting.
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 4

Some Stimmbezirke won't overlap any hospital buffer and will be missing from `stimmbezirke_covered`.
After summing the covered population per `nummer` and joining it back to `stimmbezirke`, set missing values to 0 with `tidyr::replace_na(0)`.
If a Stimmbezirk overlaps several buffers, sum its covered population across the resulting rows first and cap the resulting share at 100% with `pmin()`, just in case.
:::

### Solutions

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 1

```{r}
#| eval: false

# packages
library(dplyr)
library(readxl)
library(sf)
library(tidygeocoder)
library(tmap)


# Load address data
hospitals_cologne <- readxl::read_excel("./data/krankenhausverzeichnis.xlsx", sheet = 5, skip = 2, col_names=TRUE) |>
  dplyr::select(1:2,4:9) |>
  setNames(c("state", "district", "name", "name2", "street", "housenumber", "zip", "place")) |>
  dplyr::mutate(address = paste(street, housenumber, zip, place, "Germany")) |>
  # filter to Cologne
  dplyr::filter(grepl("K.ln", place, ignore.case = TRUE))


# Geocode
hospitals_geocoded <- hospitals_cologne |>
  tidygeocoder::geocode(address = address,
                        method = "osm")

cat("Matched:", sum(!is.na(hospitals_geocoded$lat)), "/", nrow(hospitals_geocoded), "\n")

# Convert to sf
hospitals_sf <- hospitals_geocoded |>
  dplyr::filter(!is.na(long)) |>
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326) |>
  sf::st_transform(3035)

# Map
cologne_outline <- sf::read_sf("./data/VG250_KRS.shp") |>
  dplyr::filter(GEN == "Köln") |>
  sf::st_transform(3035)

tm_shape(cologne_outline) +
  tm_polygons(col = "lightgrey") +
  tm_shape(hospitals_sf) +
  tm_dots(col = "#440154FF", size = 0.3) +
  tm_text("name", size = 0.4, ymod = 0.5)
```

```{r}
#| echo: false
#| message: false
#| warning: false

# packages
library(dplyr)
library(readxl)
library(sf)
library(tidygeocoder)
library(tmap)


# Load address data
hospitals_cologne <- readxl::read_excel("../../data/krankenhausverzeichnis.xlsx", sheet = 5, skip = 2, col_names=TRUE) |>
  dplyr::select(1:2,4:9) |>
  setNames(c("state", "district", "name", "name2", "street", "housenumber", "zip", "place")) |>
  dplyr::mutate(address = paste(street, housenumber, zip, place, "Germany")) |>
  # filter to Cologne
  dplyr::filter(grepl("K.ln", place, ignore.case = TRUE))


# Geocode
hospitals_geocoded <- hospitals_cologne |>
  tidygeocoder::geocode(address = address,
                        method = "osm")

cat("Matched:", sum(!is.na(hospitals_geocoded$lat)), "/", nrow(hospitals_geocoded), "\n")

# Convert to sf
hospitals_sf <- hospitals_geocoded |>
  dplyr::filter(!is.na(long)) |>
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326) |>
  sf::st_transform(3035)

# Map
cologne_outline <- sf::read_sf("../../data/VG250_KRS.shp") |>
  dplyr::filter(GEN == "Köln") |>
  sf::st_transform(3035)

tm_shape(cologne_outline) +
  tm_polygons(col = "lightgrey") +
  tm_shape(hospitals_sf) +
  tm_dots(col = "#440154FF", size = 0.3) +
  tm_text("name", size = 0.4, ymod = 0.5)
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 2

```{r}
library(sf)
library(dplyr)
library(tmap)

# Voting precincts
stimmbezirke <- sf::read_sf("../../data/Stimmbezirk.shp") |>
  sf::st_set_crs(25832) |>
  sf::st_transform(3035)

stimmbezirk_centroids <- sf::st_centroid(stimmbezirke)

# Point layer (hospitals from Exercise 1 — already in environment)
hospitals_sf_3035 <- sf::st_transform(hospitals_sf, 3035)

# Nearest feature index
nearest_hospital <- sf::st_nearest_feature(stimmbezirk_centroids, hospitals_sf_3035)

# Distance to nearest hospital
stimmbezirke$dist_hospital_m <- as.numeric(
  sf::st_distance(
    stimmbezirk_centroids,
    hospitals_sf_3035[nearest_hospital, ],
    by_element = TRUE
  )
)

# Map
tm_shape(stimmbezirke) +
  tm_polygons(
    fill = "dist_hospital_m",
    fill.legend = tm_legend(title = "Distance to nearest\nhospital (m)"),
    fill.scale  = tm_scale_intervals(style = "quantile", n = 5, values = "viridis"),
    col = NA
  )
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 3

```{r}
#| eval: false
# Buffer hospitals
hospital_buffers <- sf::st_buffer(hospitals_sf, dist = 1000)

# Population raster (-9 = missing)
inhabitants_cologne <- terra::rast("./data/inhabitants_cologne.tif") |>
  terra::subst(from = -9, to = NA)

# Match CRS to the raster
stimmbezirke_r     <- sf::st_transform(stimmbezirke, terra::crs(inhabitants_cologne))
hospital_buffers_r <- sf::st_transform(hospital_buffers, terra::crs(inhabitants_cologne))

# Intersect Stimmbezirke with hospital buffers
stimmbezirke_covered <- sf::st_intersection(stimmbezirke_r, hospital_buffers_r)

# Total population per Stimmbezirk
stimmbezirke_r$pop_total <- terra::extract(
  inhabitants_cologne, stimmbezirke_r, fun = sum, na.rm = TRUE
)[, 2]

# Population within the covered part(s)
stimmbezirke_covered$pop_covered <- terra::extract(
  inhabitants_cologne, stimmbezirke_covered, fun = sum, na.rm = TRUE
)[, 2]

# Sum covered population per Stimmbezirk (in case it overlaps several buffers)
pop_covered_by_district <- stimmbezirke_covered |>
  sf::st_drop_geometry() |>
  dplyr::summarise(pop_covered = sum(pop_covered), .by = nummer)

# Share of population within 1000m of a hospital
stimmbezirke <- stimmbezirke |>
  dplyr::left_join(
    stimmbezirke_r |> sf::st_drop_geometry() |> dplyr::select(nummer, pop_total),
    by = "nummer"
  ) |>
  dplyr::left_join(pop_covered_by_district, by = "nummer") |>
  dplyr::mutate(
    pop_covered   = tidyr::replace_na(pop_covered, 0),
    share_covered = pmin(pop_covered / pop_total * 100, 100)
  )

# Map
tm_shape(stimmbezirke) +
  tm_polygons(
    fill = "share_covered",
    fill.legend = tm_legend(title = "Share of population\nwithin 1000m of\na hospital (%)"),
    fill.scale  = tm_scale_intervals(style = "quantile", n = 5, values = "viridis"),
    col = NA
  )
```

```{r}
#| echo: false
#| message: false
#| warning: false
# Buffer hospitals
hospital_buffers <- sf::st_buffer(hospitals_sf, dist = 1000)

# Population raster (-9 = missing)
inhabitants_cologne <- terra::rast("../../data/inhabitants_cologne.tif") |>
  terra::subst(from = -9, to = NA)

# Match CRS to the raster
stimmbezirke_r     <- sf::st_transform(stimmbezirke, terra::crs(inhabitants_cologne))
hospital_buffers_r <- sf::st_transform(hospital_buffers, terra::crs(inhabitants_cologne))

# Intersect Stimmbezirke with hospital buffers
stimmbezirke_covered <- sf::st_intersection(stimmbezirke_r, hospital_buffers_r)

# Total population per Stimmbezirk
stimmbezirke_r$pop_total <- terra::extract(
  inhabitants_cologne, stimmbezirke_r, fun = sum, na.rm = TRUE
)[, 2]

# Population within the covered part(s)
stimmbezirke_covered$pop_covered <- terra::extract(
  inhabitants_cologne, stimmbezirke_covered, fun = sum, na.rm = TRUE
)[, 2]

# Sum covered population per Stimmbezirk (in case it overlaps several buffers)
pop_covered_by_district <- stimmbezirke_covered |>
  sf::st_drop_geometry() |>
  dplyr::summarise(pop_covered = sum(pop_covered), .by = nummer)

# Share of population within 1000m of a hospital
stimmbezirke <- stimmbezirke |>
  dplyr::left_join(
    stimmbezirke_r |> sf::st_drop_geometry() |> dplyr::select(nummer, pop_total),
    by = "nummer"
  ) |>
  dplyr::left_join(pop_covered_by_district, by = "nummer") |>
  dplyr::mutate(
    pop_covered   = tidyr::replace_na(pop_covered, 0),
    share_covered = pmin(pop_covered / pop_total * 100, 100)
  )

# Map
tm_shape(stimmbezirke) +
  tm_polygons(
    fill = "share_covered",
    fill.legend = tm_legend(title = "Share of population\nwithin 1000m of\na hospital (%)"),
    fill.scale  = tm_scale_intervals(style = "pretty", n = 5, values = "viridis"),
    col = NA
  )
```
:::
