---
title: "Exercise 5_1: Geocoding & Proximity"
subtitle: 'Geodata & Spatial Regression'
author: Anne Stroppe & Dennis Abel
format:
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
editor: visual
callout-icon: false
callout-appearance: minimal
editor_options:
  chunk_output_type: console
---

### Exercises

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 1: Geocoding

Load a number of hospitals from the hospital registry (`./data/krankenhausverzeichnis.xlsx`) or choose any other list of addresses.

1.  Geocode a set of addresses using `tidygeocoder::geocode()` with `method = "osm"`. How many were matched successfully?
2.  Convert the result to an `sf` point layer (CRS 4326), then transform to CRS 3035.
3.  Map the geocoded hospitals.
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 1

```{r}
#| eval: false

# load hospital data
hospitals_cologne <- readxl::read_excel("data/krankenhausverzeichnis.xlsx", sheet = 5, skip = 2, col_names=TRUE) |>
  dplyr::select(1:2,4:9) |>
  setNames(c("state", "district", "name", "name2", "street", "housenumber", "zip", "place")) |>
  dplyr::mutate(address = paste(street, housenumber, zip, place, "Germany")) 
```
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 2

Use `sf::st_as_sf(coords = c("long", "lat"), crs = 4326)` to convert.
Note the order: longitude first, then latitude.
Check for `NA` coordinates before converting with `dplyr::filter(!is.na(long))`.
:::

:::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 2: Distance & Nearest Neighbour

Usinga point dataset of your choice and a multipolygon or raster layer, compute the distance from each centroid to the nearest point feature.
Map the result.

**Choose your point dataset — pick whichever fits your interests:**

::: panel-tabset
### Option A: Hospitals (from Exercise 1)

Use the hospitals you already geocoded.

```{r}
#| eval: false
# Use the hospitals_sf object from Exercise 1
hospitals_sf_3035 <- sf::st_transform(hospitals_cologne, 4326)
```

### Option B: NRW train stops (data folder)

Use the train stop data already in the `./data` folder.
Note: CRS is missing, set it manually.

```{r}
#| eval: false
train_stops <- sf::read_sf("./data/nrw_pt_osmtrainstops.shp") |>
  sf::st_set_crs(25832) |>
  sf::st_transform(4326)   
```

### Option C: Your own OSM feature

Query any feature from OpenStreetMap using the `osmdata` package (as in Session 4).
Examples: pharmacies, schools, supermarkets, parks, bus stops.

```{r}
#| eval: false
library(osmdata)

your_features <- osmdata::getbb("Köln") |>
  osmdata::opq(timeout = 25) |>
  osmdata::add_osm_feature(
    key   = "amenity",       # change key/value to your feature of interest
    value = "pharmacy"       # e.g. "school", "supermarket", "post_office"
  ) |>
  osmdata::osmdata_sf() |>
  _$osm_points 
```
:::
::::


::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

**Steps for all options:**

1.  Load `Stimmbezirk.shp`, set CRS to 3035, compute centroids with `sf::st_centroid()`
2.  Find the nearest feature for each centroid using `sf::st_nearest_feature()`
3.  Compute distances with `sf::st_distance(..., by_element = TRUE)`
4.  Map the result: Which precincts have the worst access?
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

Remember: `st_nearest_feature()` returns an *index*, not a distance.
Use that index to subset your point layer before calling `st_distance(..., by_element = TRUE)`.
Convert the result to numeric metres with `as.numeric()`.
:::

:::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 3: Buffers & Hospital Coverage

So far, we have measured hospital accessibility as the straight-line distance from each polling district's centroid to the nearest hospital.
This is a useful first approximation but how convincing is it really?
Consider what a centroid represents: a single point summarising an entire area.
A polling district might have its centroid close to a hospital, yet most of its residents could live at the far end of the district, far from any facility.
The centroid distance measure implicitly assumes that all residents live at that one point but this assumption becomes increasingly problematic as districts grow larger or more irregularly shaped.

Instead, we are interes in the question: **What share of each polling district's population lives within reach of a hospital?**

In Session 4, we used `sf::st_buffer()` to create circular buffers around points and `terra::extract()` to summarize raster values within them.
Let's apply the same idea here: instead of measuring *distance* to the nearest hospital.

::: panel-tabset
# Workflow
0. Load necessary data.
1.  Create a 1,000m buffer around each hospital in `hospitals_sf` using `sf::st_buffer()`.
2.  Intersect the Cologne `stimmbezirke` (from Exercise 2) with these buffers to get the part of each precinct that lies within 1,000m of a hospital.
3.  Load `./data/inhabitants_cologne.tif` and use `terra::extract()` to get (a) the total population per Stimmbezirk and (b) the population within the "covered" part(s).
4.  Compute the share (in %) of each Stimmbezirk's population that lives within 1,000m of a hospital, and map the result.
:::
::::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Create and load data
```{r}
#| eval: false


# Voting precincts
stimmbezirke <- sf::read_sf("./data/Stimmbezirk.shp") |>
  sf::st_set_crs(25832) |>
  sf::st_transform(3035)

# Population raster (-9 = missing)
inhabitants_cologne <- terra::rast("../../data/inhabitants_cologne.tif") |>
  terra::subst(from = -9, to = NA)

# Geocoded Hospitals
# Load address data
hospitals_cologne <- readxl::read_excel("./data/krankenhausverzeichnis.xlsx", sheet = 5, skip = 2, col_names=TRUE) |>
  dplyr::select(1:2,4:9) |>
  setNames(c("state", "district", "name", "name2", "street", "housenumber", "zip", "place")) |>
  dplyr::mutate(address = paste(street, housenumber, zip, place, "Germany")) |>
  # filter to Cologne
  dplyr::filter(grepl("K.ln", place, ignore.case = TRUE))


# Geocode
hospitals_geocoded <- hospitals_cologne |>
  tidygeocoder::geocode(address = address,
                        method = "osm")

# Convert to sf
hospitals_sf <- hospitals_geocoded |>
  dplyr::filter(!is.na(long)) |>
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326) |>
  sf::st_transform(3035)


# Population raster (-9 = missing)
inhabitants_cologne <- terra::rast("./data/inhabitants_cologne.tif") 


```
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 1

```{r}
#| eval: false

hospital_buffers <- sf::st_buffer(hospitals_sf, dist = 1000)
```
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 2

`sf::st_intersection(x, y)` returns only the parts of `x`'s geometries that overlap with `y`, keeping `x`'s attributes — unlike `sf::st_join()`, the geometries themselves get cut.

```{r}
#| eval: false
stimmbezirke_covered <- sf::st_intersection(stimmbezirke, hospital_buffers)
```

Reminder: If a Stimmbezirk overlaps more than one hospital buffer, it appears as multiple rows in the result.
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 3

Make sure `stimmbezirke` and `hospital_buffers` are transformed to the raster's CRS (`terra::crs(inhabitants_cologne)`) before extracting.
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 4

Some Stimmbezirke won't overlap any hospital buffer and will be missing from `stimmbezirke_covered`.
After summing the covered population per `nummer` and joining it back to `stimmbezirke`, set missing values to 0 with `tidyr::replace_na(0)`.
If a Stimmbezirk overlaps several buffers, sum its covered population across the resulting rows first and cap the resulting share at 100% with `pmin()`, just in case.
:::

### Solutions

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 1

```{r}
#| eval: false

# packages
library(dplyr)
library(readxl)
library(sf)
library(tidygeocoder)
library(ggplot2)


# Load address data
hospitals_bremen <- readxl::read_excel("./data/krankenhausverzeichnis.xlsx", sheet = 5, skip = 2, col_names=TRUE) |>
  dplyr::select(1:2,4:9) |>
  setNames(c("state", "district", "name", "name2", "street", "housenumber", "zip", "place")) |>
  dplyr::mutate(address = paste(street, housenumber, zip, place, "Germany")) |>
  # filter to Cologne
  dplyr::filter(grepl("Bremen", place, ignore.case = TRUE))


# Geocode
hospitals_geocoded <- hospitals_bremen |>
  tidygeocoder::geocode(address = address,
                        method = "osm")

cat("Matched:", sum(!is.na(hospitals_geocoded$lat)), "/", nrow(hospitals_geocoded), "\n")

# Convert to sf
hospitals_sf <- hospitals_geocoded |>
  dplyr::filter(!is.na(long)) |>
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326) |>
  sf::st_transform(3035)

# Map
bremen_outline <- sf::read_sf("../../data/VG250_KRS.shp") |>
  dplyr::filter(GEN == "Bremen") |>
  sf::st_transform(3035)


ggplot() +
  geom_sf(
    data  = bremen_outline,
    fill  = "lightgrey",
    color = NA
  ) +
  geom_sf(
    data  = hospitals_sf,
    color = "#440154FF",
    size  = 0.3
  ) +
  ggrepel::geom_text_repel(
    data        = hospitals_sf,
    aes(label   = name,
        geometry = geometry),
    stat        = "sf_coordinates",
    size        = 2,          
    nudge_y     = 0.005,      
    min.segment.length = 0,   
    color       = "black"
  ) +
  theme_void()
```

```{r solf-one}
#| echo: false
#| message: false
#| warning: false

# packages
library(dplyr)
library(readxl)
library(sf)
library(ggplot2)

# Load address data
hospitals_bremen <- readxl::read_excel("../../data/krankenhausverzeichnis.xlsx", sheet = 5, skip = 2, col_names=TRUE) |>
  dplyr::select(1:2,4:9) |>
  setNames(c("state", "district", "name", "name2", "street", "housenumber", "zip", "place")) |>
  dplyr::mutate(address = paste(street, housenumber, zip, place, "Germany")) |>
  # filter to Cologne
  dplyr::filter(grepl("Bremen", place, ignore.case = TRUE))


# Geocode
hospitals_geocoded <- hospitals_bremen |>
  tidygeocoder::geocode(address = address,
                        method = "osm")

cat("Matched:", sum(!is.na(hospitals_geocoded$lat)), "/", nrow(hospitals_geocoded), "\n")

# Convert to sf
hospitals_sf <- hospitals_geocoded |>
  dplyr::filter(!is.na(long)) |>
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326) |>
  sf::st_transform(3035)

# Map
bremen_outline <- sf::read_sf("../../data/VG250_KRS.shp") |>
  dplyr::filter(GEN == "Bremen") |>
  sf::st_transform(3035)


ggplot() +
  geom_sf(
    data  = bremen_outline,
    fill  = "lightgrey",
    color = NA
  ) +
  geom_sf(
    data  = hospitals_sf,
    color = "#440154FF",
    size  = 0.3
  ) +
  ggrepel::geom_text_repel(
    data        = hospitals_sf,
    aes(label   = name,
        geometry = geometry),
    stat        = "sf_coordinates",
    size        = 2,          
    nudge_y     = 0.005,      
    min.segment.length = 0,   
    color       = "black"
  ) +
  theme_void()
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 2

```{r sol-two}
library(sf)
library(dplyr)
library(ggplot2)

# Point layer (hospitals from Exercise 1 — already in environment)
hospitals_sf_4326 <- sf::st_transform(hospitals_sf, 4326)

# Query OSM for administrative boundaries

# admin_level meanings for Germany:
#   4  = Bundesland
#   5  = Regierungsbezirk
#   6  = Landkreis / kreisfreie Stadt
#   7  = Amt / Verbandsgemeinde
#   8  = Gemeinde  ← most common for municipality level
#   9  = Gemeindeteil / Ortsteil
#
# Adjust admin_level to your needs:

osm_query <- osmdata::opq(bbox = sf::st_bbox(hospitals_sf_4326)) |>
  osmdata::add_osm_feature(
    key   = "boundary",
    value = "administrative"
  ) |>
  osmdata::add_osm_feature(
    key   = "admin_level",
    value = "9"          # ← change this to your target level
  )

osm_result <- osmdata::osmdata_sf(osm_query)

#  Extract the multipolygon layer
admin_polygons <- osm_result$osm_multipolygons

# Quick check on what came back
dplyr::glimpse(admin_polygons)

# Centroids
admin_centroids <- sf::st_centroid(admin_polygons)

# Nearest feature index
nearest_hospital <- sf::st_nearest_feature(admin_centroids, hospitals_sf_4326)

# Distance to nearest hospital
admin_centroids$dist_to_hospital <- as.numeric(
  sf::st_distance(
    admin_centroids,
    hospitals_sf_4326[nearest_hospital, ],
    by_element = TRUE
  )
)

admin_polygons$dist_to_hospital_m <- as.numeric(admin_centroids$dist_to_hospital)

# Map
ggplot() +
  geom_sf(
    data = admin_polygons,
    aes(fill = dist_to_hospital_m)   # no quotes
  ) +
  scale_fill_viridis_c(
    name   = "Distance to\nnearest hospital (m)",
    option = "viridis"            # or "magma", "plasma", "inferno", "mako"
  ) +
  theme_void()
```
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 3

```{r sol-three}
#| eval: false

# Define mmissings
inhabitants_cologne <- 
  inhabitants_cologne |>
  terra::subst(from = -9, to = NA)

# Buffer hospitals
hospital_buffers <- sf::st_buffer(hospitals_sf, dist = 1000)

# Match CRS to the raster
stimmbezirke_r     <- sf::st_transform(stimmbezirke, terra::crs(inhabitants_cologne))
hospital_buffers_r <- sf::st_transform(hospital_buffers, terra::crs(inhabitants_cologne))

# Intersect Stimmbezirke with hospital buffers
stimmbezirke_covered <- sf::st_intersection(stimmbezirke_r, hospital_buffers_r)

# Total population per Stimmbezirk
stimmbezirke_r$pop_total <- terra::extract(
  inhabitants_cologne, stimmbezirke_r, fun = sum, na.rm = TRUE
)[, 2]

# Population within the covered part(s)
stimmbezirke_covered$pop_covered <- terra::extract(
  inhabitants_cologne, stimmbezirke_covered, fun = sum, na.rm = TRUE
)[, 2]

# Sum covered population per Stimmbezirk (in case it overlaps several buffers)
pop_covered_by_district <- stimmbezirke_covered |>
  sf::st_drop_geometry() |>
  dplyr::summarise(pop_covered = sum(pop_covered), .by = nummer)

# Share of population within 1000m of a hospital
stimmbezirke <- stimmbezirke |>
  dplyr::left_join(
    stimmbezirke_r |> sf::st_drop_geometry() |> dplyr::select(nummer, pop_total),
    by = "nummer"
  ) |>
  dplyr::left_join(pop_covered_by_district, by = "nummer") |>
  dplyr::mutate(
    pop_covered   = tidyr::replace_na(pop_covered, 0),
    share_covered = pmin(pop_covered / pop_total * 100, 100)
  )

# Map
ggplot() +
  geom_sf(
    data  = stimmbezirke,
    aes(fill = share_covered),
    color = NA
  ) +
  scale_fill_viridis_b(
    name     = "Share of population\nwithin 1000m of\na hospital (%)",
    n.breaks = 5,
    option   = "viridis"
  ) +
  theme_void()
```

```{r}
#| echo: false
#| message: false
#| warning: false

# Voting precincts
stimmbezirke <- sf::read_sf("../../data/Stimmbezirk.shp") |>
  sf::st_set_crs(25832) |>
  sf::st_transform(3035)

# Population raster (-9 = missing)
inhabitants_cologne <- terra::rast("../../data/inhabitants_cologne.tif") |>
  terra::subst(from = -9, to = NA)

# Geocoded Hospitals
# Load address data
hospitals_cologne <- readxl::read_excel("../../data/krankenhausverzeichnis.xlsx", sheet = 5, skip = 2, col_names=TRUE) |>
  dplyr::select(1:2,4:9) |>
  setNames(c("state", "district", "name", "name2", "street", "housenumber", "zip", "place")) |>
  dplyr::mutate(address = paste(street, housenumber, zip, place, "Germany")) |>
  # filter to Cologne
  dplyr::filter(grepl("K.ln", place, ignore.case = TRUE))


# Geocode
hospitals_geocoded <- hospitals_cologne |>
  tidygeocoder::geocode(address = address,
                        method = "osm")

# Convert to sf
hospitals_sf <- hospitals_geocoded |>
  dplyr::filter(!is.na(long)) |>
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326) |>
  sf::st_transform(3035)


# Buffer hospitals
hospital_buffers <- sf::st_buffer(hospitals_sf, dist = 1000)

# Match CRS to the raster
stimmbezirke_r     <- sf::st_transform(stimmbezirke, terra::crs(inhabitants_cologne))
hospital_buffers_r <- sf::st_transform(hospital_buffers, terra::crs(inhabitants_cologne))

# Intersect Stimmbezirke with hospital buffers
stimmbezirke_covered <- sf::st_intersection(stimmbezirke_r, hospital_buffers_r)

# Total population per Stimmbezirk
stimmbezirke_r$pop_total <- terra::extract(
  inhabitants_cologne, stimmbezirke_r, fun = sum, na.rm = TRUE
)[, 2]

# Population within the covered part(s)
stimmbezirke_covered$pop_covered <- terra::extract(
  inhabitants_cologne, stimmbezirke_covered, fun = sum, na.rm = TRUE
)[, 2]

# Sum covered population per Stimmbezirk (in case it overlaps several buffers)
pop_covered_by_district <- stimmbezirke_covered |>
  sf::st_drop_geometry() |>
  dplyr::summarise(pop_covered = sum(pop_covered), .by = nummer)

# Share of population within 1000m of a hospital
stimmbezirke <- stimmbezirke |>
  dplyr::left_join(
    stimmbezirke_r |> sf::st_drop_geometry() |> dplyr::select(nummer, pop_total),
    by = "nummer"
  ) |>
  dplyr::left_join(pop_covered_by_district, by = "nummer") |>
  dplyr::mutate(
    pop_covered   = tidyr::replace_na(pop_covered, 0),
    share_covered = pmin(pop_covered / pop_total * 100, 100)
  )

# Map
ggplot() +
  geom_sf(
    data  = stimmbezirke,
    aes(fill = share_covered),
    color = NA
  ) +
  scale_fill_viridis_b(
    name     = "Share of population\nwithin 1000m of\na hospital (%)",
    n.breaks = 5,
    option   = "viridis"
  ) +
  theme_void()
```
:::
