---
title: "Exercise 5_2: Incongruent Units"
subtitle: 'Geodata & Spatial Regression'
author: Anne-Kathrin Stroppe & Dennis Abel
format:
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
editor: visual
callout-icon: false
callout-appearance: minimal
editor_options:
  chunk_output_type: console
---

```{r}
#| echo: false
#| include: false
library(sf)
library(dplyr)
library(tmap)
```

### Exercises

In this exercise you practice the three methods for linking survey data across incongruent spatial units — the same workflow as the [AreaMatch tool](https://kodaqs-toolbox.gesis.org/github.com/StefanJuenger/zipmatching/index/). You will work with German ZIP code areas (`plz_gebiete.gpkg`), municipality boundaries (`gemeindegrenzen_2022.gpkg`), and simulated survey data (`simulated_survey_data.rds`).

---

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 1: Geometric Intersection

Load the ZIP code areas and municipality boundaries. Select a small region to work with (e.g., filter to one federal state using the `ags` column prefix, or use `dplyr::slice(1:100)`).

1. Run `sf::st_intersection()` on your ZIP code subset and the municipalities. How many features does the result have compared to the input?
2. Add an `area_m2` column with the area of each intersected piece.
3. For each ZIP code (`zip_code`), compute the share of its area that falls within each municipality. Which ZIP codes straddle more than one municipality?
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

Use `sf::st_make_valid()` on both layers before intersecting. Use `dplyr::mutate(share = area_m2 / sum(area_m2), .by = zip_code)` to compute the proportional overlap per ZIP code.
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution

```{r}
zip_codes <- sf::st_read("../../data/plz_gebiete.gpkg", quiet = TRUE) |>
  dplyr::select(zip_code = plz, inhabitants_zip = einwohner)

municipalities <- sf::st_read("../../data/gemeindegrenzen_2022.gpkg", quiet = TRUE) |>
  dplyr::select(ags = AGS, inhabitants_mun = EWZ, area_mun = KFL)

# Work with a small subset
zip_sub  <- sf::st_make_valid(zip_codes[1:80, ])
mun_sub  <- sf::st_make_valid(municipalities)

intersection <- sf::st_intersection(zip_sub, mun_sub)
intersection$area_m2 <- as.numeric(sf::st_area(intersection))

cat("Input ZIP codes:", nrow(zip_sub), "\n")
cat("After intersection:", nrow(intersection), "\n")

# Overlap shares
overlap_shares <- intersection |>
  sf::st_drop_geometry() |>
  dplyr::mutate(share = area_m2 / sum(area_m2), .by = zip_code)

# ZIP codes straddling more than one municipality
split_zips <- overlap_shares |>
  dplyr::filter(dplyr::n() > 1, .by = zip_code)
cat("ZIP codes split across municipalities:", dplyr::n_distinct(split_zips$zip_code), "\n")
head(split_zips |> dplyr::select(zip_code, ags, area_m2, share), 6)
```
:::

---

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 2: Three Linking Methods

Load the simulated survey data (`simulated_survey_data.rds`). Filter to valid ZIP codes (non-missing, 5 characters, present in the `zip_codes` layer) and join to the ZIP code geometries.

Apply all three methods to assign **municipality population** (`inhabitants_mun`) to each survey respondent's ZIP code:

1. **Centroid matching** — `sf::st_point_on_surface()` + `sf::st_join()`
2. **Areal matching** — `sf::st_join(..., largest = TRUE)`
3. **Areal interpolation** — `sf::st_interpolate_aw(..., extensive = FALSE)`

Compare the resulting `inhabitants_mun` values across methods for the same ZIP codes. Do they differ?
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

For `sf::st_interpolate_aw()`: the first argument is the **source** (municipalities with the attribute), the second is the **target** (ZIP code geometries). Use `extensive = FALSE` for intensive variables (rates, densities) and `extensive = TRUE` for counts.
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution

```{r}
survey <- readRDS("../../data/simulated_survey_data.rds")

# Valid ZIP codes joined to geometries
zip_codes_valid <- survey |>
  dplyr::filter(!is.na(zip_code), nchar(zip_code) == 5,
                zip_code %in% zip_codes$zip_code) |>
  dplyr::left_join(zip_codes, by = "zip_code") |>
  sf::st_as_sf()

cat("Valid survey responses:", nrow(zip_codes_valid), "\n")

# Method 1: Centroid matching
centroid_matched <- zip_codes_valid |>
  sf::st_point_on_surface() |>
  sf::st_join(municipalities) |>
  dplyr::select(zip_code, inhabitants_zip, cent_inh_mun = inhabitants_mun) |>
  dplyr::distinct()

# Method 2: Areal matching
areal_matched <- zip_codes_valid |>
  sf::st_join(municipalities, left = TRUE, largest = TRUE) |>
  dplyr::select(zip_code, inhabitants_zip, areal_inh_mun = inhabitants_mun) |>
  dplyr::distinct()

# Method 3: Areal interpolation
interp_matched <- sf::st_interpolate_aw(
  municipalities["inhabitants_mun"],
  zip_codes_valid,
  extensive = FALSE
) |>
  dplyr::bind_cols(
    zip_codes_valid |> sf::st_drop_geometry() |>
      dplyr::select(zip_code, inhabitants_zip)
  ) |>
  dplyr::select(zip_code, interp_inh_mun = inhabitants_mun) |>
  dplyr::distinct()

# Compare
comparison <- centroid_matched |>
  sf::st_drop_geometry() |>
  dplyr::left_join(areal_matched |> sf::st_drop_geometry(), by = c("zip_code", "inhabitants_zip")) |>
  dplyr::left_join(interp_matched |> sf::st_drop_geometry(), by = "zip_code")

head(comparison, 8)
```
:::

---

::: callout-note
## `r emo::ji("weight_lifting_man")` Bonus Exercise: Which method gives the best result?

The AreaMatch tool evaluates method accuracy by comparing each method's output against "ground truth" — the actual municipality of simulated respondent points within each ZIP code.

Explore the [AreaMatch tool](https://kodaqs-toolbox.gesis.org/github.com/StefanJuenger/zipmatching/index/) and run its full evaluation workflow. Which method has the smallest error? Under what conditions does areal interpolation outperform centroid or areal matching?
:::
