---
title: "Exercise 6: Spatial Joins"
subtitle: 'Geodata & Spatial Regression'
author: Anne-Kathrin Stroppe & Dennis Abel
format: 
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
editor: visual
callout-icon: false
callout-appearance: minimal
editor_options: 
  chunk_output_type: console
---

### Exercises

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 1: Geocoding

Load a number of hospitals from the hospital registry (`./data/krankenhausverzeichnis.xlsx`) or choose any other list of adresses.

1.  Geocode a set of adresses using `tidygeocoder::geocode()` with `method = "osm"`. How many were matched successfully?
3.  Convert the result to an `sf` point layer (CRS 4326), then transform to CRS 3035.
4.  Map the geocoded hospitals .
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 1

```{r}
#| eval: false

# load hospital data
hospitals_cologne <- readxl::read_excel("data/krankenhausverzeichnis.xlsx", sheet = 5, skip = 2, col_names=TRUE) |> 
  dplyr::select(1:2,4:9) |>
  setNames(c("state", "district", "name", "name2", "street", "housenumber", "zip", "place")) |> 
  dplyr::mutate(address = paste(strasse, hausnr, plz, ort, "Germany"))   |> 
  # filter bei 
  dplyr::filter(grepl("K.ln", ort, ignore.case = TRUE)) 

```

:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip 2

Use `sf::st_as_sf(coords = c("long", "lat"), crs = 4326)` to convert. Note the order: longitude first, then latitude.
Check for `NA` coordinates before converting with `dplyr::filter(!is.na(long))`.
:::



:::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 2: Distance & Nearest Neighbour

Using the Cologne voting precincts (`Stimmbezirk.shp`) and a point dataset of your choice, compute the distance from each precinct centroid to the nearest point feature.
Map the result.

**Choose your point dataset — pick whichever fits your interests:**

::: panel-tabset
### 🏥 Option A: Hospitals (from Exercise 0)

Use the hospitals you already geocoded.
This continues naturally from Exercise 0.

```{r}
#| eval: false
# Use the hospitals_sf object from Exercise 0
hospitals_sf_3035 <- sf::st_transform(hospitals_sf, 3035)
```

### 🚉 Option B: NRW train stops (pre-loaded)

Use the train stop data already in the `./data` folder.
Note: CRS is missing — set it manually.

```{r}
#| eval: false
train_stops <- sf::read_sf("./data/nrw_pt_osmtrainstops.shp") |>
  sf::st_set_crs(25832) |>
  sf::st_transform(3035)   # coordinates already in LAEA Europe
```

### 🗺️ Option C: Your own OSM feature

Query any feature from OpenStreetMap using the `osmdata` package (as in Session 4).
Examples: pharmacies, schools, supermarkets, parks, bus stops.

```{r}
#| eval: false
library(osmdata)

your_features <- osmdata::getbb("Köln") |>
  osmdata::opq(timeout = 25) |>
  osmdata::add_osm_feature(
    key   = "amenity",       # change key/value to your feature of interest
    value = "pharmacy"       # e.g. "school", "supermarket", "post_office"
  ) |>
  osmdata::osmdata_sf() |>
  _$osm_points |>
  sf::st_transform(3035)
```
:::

**Steps for all options:**

1.  Load `Stimmbezirk.shp`, set CRS to 3035, compute centroids with `sf::st_centroid()`
2.  Find the nearest feature for each centroid using `sf::st_nearest_feature()`
3.  Compute distances with `sf::st_distance(..., by_element = TRUE)`
4.  Map the result — which precincts have the worst access?
::::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

Remember: `st_nearest_feature()` returns an *index*, not a distance.
Use that index to subset your point layer before calling `st_distance(..., by_element = TRUE)`.
Convert the result to numeric metres with `as.numeric()`.
:::

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution (Option A: hospitals)

```{r}
library(sf); library(dplyr); library(tmap)

# Voting precincts
stimmbezirke <- sf::read_sf("../../data/Stimmbezirk.shp") |>
  sf::st_set_crs(25832) |>
  sf::st_transform(3035)

stimmbezirk_centroids <- sf::st_centroid(stimmbezirke)

# Point layer (hospitals from Exercise 0 — already in environment)
hospitals_sf_3035 <- sf::st_transform(hospitals_sf, 3035)

# Nearest feature index
nearest_idx <- sf::st_nearest_feature(stimmbezirk_centroids, hospitals_sf_3035)

# Distance to nearest hospital
stimmbezirke$dist_hospital_m <- as.numeric(
  sf::st_distance(
    stimmbezirk_centroids,
    hospitals_sf_3035[nearest_idx, ],
    by_element = TRUE
  )
)

# Map
tm_shape(stimmbezirke) +
  tm_polygons(
    fill = "dist_hospital_m",
    fill.legend = tm_legend(title = "Distance to nearest\nhospital (m)"),
    fill.scale  = tm_scale_intervals(style = "quantile", n = 5, values = "viridis"),
    col = NA
  )
```
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 2

We may not always rely on georeferenced data collected in the whole area of a country.
Thus, when exploiting geospatial data for linking, one exercise is to subset the data.
This could be a subset of the charging station data used in the lecture.
Let's say we conducted a survey in the German federal state North-Rhine Westphalia (NRW).

Subset the charger data file yourself by relying on the spatial information of the file `charging_points_ger.csv` and a polygon of NRW.
There are two ways to achieve this.
How many chargers are located within NRW?
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

You need two datasets for that: the point layer `charging_points_ger.csv` (remember to adjust the crs) in the `./data` folder and polygons of NRW.
For the latter, you can again use the `osmdata` syntax or one of the shapefiles in the `./data/folder` you filter for NRW.
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

You can explore two functions: `sf::st_join` and `sf::st_intersection()`.
The default of `sf::st_join()` will leave you with a 'left-join' and returns a data object with all chargers and matching district information for those located within NRW.
You can reset the option to perform an 'inner-join' and keep only the observation that lay within the predefined area (`sf::st_join(x , y, join  = "", left = FALSE)`).
:::

::: callout-note
## `r emo::ji("weight_lifting_man")` Exercise 3

Did the operationalization of charger accessibility convince you?
There may be other ways, such as the proportion of residents with max.
1000 m linear distance to the nearest charger.

You may not know this, but we have everything we need to create this indicator on a smaller scale as well: charger data, Census data, and our synthetic georeferenced data!
So, what is the mean share of residents with max.
1000m linear distance to the nearest train station in a 5km neighborhood of our fake respondents?

You can run the code below to load all the data you need.
:::

```{r}
#| eval: false
#| message: false
#| warning: false
# packages
library(dplyr)
library(osmdata)
library(sf)
library(terra)

# load charger
charger_ger <- 
  # Read charging station points datae
  readr::read_delim("./data/charging_points_ger.csv", delim = ";") |> 
  # Filter out rows with missing longitude or latitude
  dplyr::filter(!is.na(longitude) & !is.na(latitude)) |> 
  # Convert data frame to sf object
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# nrw polygon
nrw <-
  osmdata::getbb(
    "Nordrhein-Westfalen", 
    format_out = "sf_polygon"
  ) |>  
  _$multipolygon

# charger nrw
charger_nrw <- 
  charger_ger |> 
  # Subset point data to sampling area
  sf::st_intersection(nrw)

# Census data
census_inhabitants <- terra::rast("./data/census_inhabitants.tiff") 

# crs transformations
nrw <- sf::st_transform(nrw, 3035)
charger_nrw <- sf::st_transform(charger_nrw, 3035)

# Synthetic georeferenced data
synthetic_data <- 
  readRDS("./data/synthetic_survey_coordinates.rds") |> 
  sf::st_intersection(nrw)
```

```{r}
#| echo: false
#| message: false
#| warning: false
# packages
library(dplyr)
library(osmdata)
library(sf)
library(terra)

# load charger
charger_ger <- 
  # Read charging station points datae
  readr::read_delim("../../data/charging_points_ger.csv", delim = ";") |> 
  # Filter out rows with missing longitude or latitude
  dplyr::filter(!is.na(longitude) & !is.na(latitude)) |> 
  # Convert data frame to sf object
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# nrw polygon
nrw <-
  osmdata::getbb(
    "Nordrhein-Westfalen", 
    format_out = "sf_polygon"
  ) |>  
  _$multipolygon

# charger nrw
charger_nrw <- 
  charger_ger |> 
  # Subset point data to sampling area
  sf::st_intersection(nrw)

# Census data
census_inhabitants <- terra::rast("../../data/census_inhabitants.tiff")

# crs transformations
nrw <- sf::st_transform(nrw, 3035)
charger_nrw <- sf::st_transform(charger_nrw, 3035)

# Synthetic georeferenced data
synthetic_data <- 
  readRDS("../../data/synthetic_survey_coordinates.rds") |> 
  sf::st_intersection(nrw)
```

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

As always, there are several ways to do this.
We suggest the following steps:

1.  Create a point layer with the centroids of all grids in NRW based on the census population layer.

2.  Calculate the distance to the next charger station for each grid.

3.  Create a column that equals 0 if the distance to the next charger is \>1000m and contains the number of inhabitants if \<1000.

4.  Rasterize the sf data object to receive two raster objects: number of inhabitants and number of inhabitants with a max 1000m distance to the next train station.

5.  Calculate the mean for the 5km buffer for each "respondent" for each raster.

6.  Calculate the share.
:::

::: {.callout-caution collapse="true"}
## `r emo::ji("bulb")` Tip

In the add-on slides of the spatial wrangling session, we introduced and gave some information on how to transform the raster to points (and back).
To get a `sf` point layer for the raster object, you can use `terra::as.points() |> sf::st_as_sf()`.
To rasterize the object, you can use `terra::rasterize(input_points, target_raster, input_raster, field = "target_variable", fun = "mean", background = 0)`.
:::

### Solutions

::: {.callout-tip collapse="true"}
## `r emo::ji("white_check_mark")` Solution 1

```{r}
#| eval: false

# packages
library(dplyr)
library(readxl)
library(sf)
library(tidygeocoder)
library(tmap)


# Load address data
bremen_hospitals <- readxl::read_excel("./data/krankenhausverzeichnis.xlsx", sheet = 5, skip = 2, col_names=TRUE) |> 
  dplyr::select(1:2,4:9) |>
  setNames(c("state", "district", "name", "name2", "street", "housenumber", "zip", "place")) |> 
  # filter bei 
  dplyr::filter(state==04)  


# Geocode
hospitals_geocoded <- bremen_hospitals |>
  tidygeocoder::geocode(address = address, 
                        method = "osm")

cat("Matched:", sum(!is.na(hospitals_geocoded$lat)), "/", nrow(hospitals_geocoded), "\n")

# Convert to sf
hospitals_sf <- hospitals_geocoded |>
  dplyr::filter(!is.na(long)) |>
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326) |>
  sf::st_transform(3035)

# Map
bremen <- sf::read_sf("../../data/VG250_KRS.shp") |>
  dplyr::filter(GEN == "Bremen") |>
  sf::st_transform(3035)

tm_shape(bremen) +
  tm_polygons(col = "lightgrey") +
  tm_shape(hospitals_sf) +
  tm_dots(col = "#440154FF", size = 0.3) +
  tm_text("name", size = 0.4, ymod = 0.5)
```


```{r}
#| echo: false
#| message: false
#| warning: false

# packages
library(dplyr)
library(readxl)
library(sf)
library(tidygeocoder)
library(tmap)


# Load address data
bremen_hospitals <- readxl::read_excel("../../data/krankenhausverzeichnis.xlsx", sheet = 5, skip = 2, col_names=TRUE) |> 
  dplyr::select(1:2,4:9) |>
  setNames(c("state", "district", "name", "name2", "street", "housenumber", "zip", "place")) |> 
  # filter bei 
  dplyr::filter(state==04)  


# Geocode
hospitals_geocoded <- bremen_hospitals |>
  tidygeocoder::geocode(address = address, 
                        method = "osm")

cat("Matched:", sum(!is.na(hospitals_geocoded$lat)), "/", nrow(hospitals_geocoded), "\n")

# Convert to sf
hospitals_sf <- hospitals_geocoded |>
  dplyr::filter(!is.na(long)) |>
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326) |>
  sf::st_transform(3035)

# Map
bremen <- sf::read_sf("../../data/VG250_KRS.shp") |>
  dplyr::filter(GEN == "Bremen") |>
  sf::st_transform(3035)

tm_shape(bremen) +
  tm_polygons(col = "lightgrey") +
  tm_shape(hospitals_sf) +
  tm_dots(col = "#440154FF", size = 0.3) +
  tm_text("name", size = 0.4, ymod = 0.5)
```
:::

